### 2024-09-04 teaching/math240f24/draft.tex
Fixed formatting
Lecture 3 september

Lastly, we define the {\it symmetric difference} $A \symdiff B$ of two sets $A$ and $B$ to be the set of all
elements that are either in $A$ or in $B$ {\it but not both}.
Invoking the above example one last time, we
have $A\symdiff B = \{2,3,4,5\}$.
We can express it as using unions, intersections,
and complements as
\edef\eqsymdiffdef{\the\eqcount}
$$A \symdiff B = (A\cup B) \cap \bar{A \cap B}.\adveq$$
To practise using all the different operations we just learned,
convince yourself that the following are three more valid ways to express the symmetric difference:
$$A\symdiff B = (A\cup B)\setminus (A\cap B) = (A\setminus B) \cup (B\setminus A) =
\bar{\bar{A\cup B} \cup (A\cap B)}$$
\datestamp{2}{} More set identities abound. We state the following proposition without proof; you should try going
though this list and convincing yourself that each identity holds, for all sets $A$, $B$, and $C$.
(This is a great way of practising proofs by double inclusion.)

@@

These laws have names, some of which we will use more often than others:
(i) is called the identity law, (ii) the domination law, (iii) the idempotent law,
(iv) the law of double negation, (v) the commutative law, (vi) the associative law,
(vii) the distributive law, (viii) the absorption law, and (ix) the complement law.
\bye

@@

\medskip\boldlabel De Morgan's laws. There are two important laws relating complements with union and intersection.
We shall state them as a proposition, this time giving a proof (of one of them).

@@

To actually prove that $2^{A\cup B} \subseteq 2^A \cup 2^B$ doesn't hold in general, we need to find
a {\it counterexample}. That is, we need to construct sets $A$ and $B$ such that the statement is false.
In this case, we can let $A = \{1,2\}$, $B = \{3,4\}$, so that $A\cup B = \{1,2,3,4\}$.
Then the set $\{1,3\}$ is a subset of $A\cup B$ but is not a subset of $A$ and it is not a subset of $B$.
In other words, $\{1,3\}\in 2^{A\cup B}$ but $\{1,3\} \notin 2^A\cup 2^B$, proving that
$2^{A\cup B} \not\subseteq 2^A \cup 2^B$.

@@

A {\it function} $f$ from a set $A$ to a set $B$ is a subset $f\subseteq A\times B$ such that
for every $a\in A$, there is exactly one $b\in B$ such that $(a,b)\in f$. If $(a,b)\in f$,
we write $f(a) = b$. The set $A$ is called the {\it domain} and the set $B$ is the {\it codomain}.
The notation $f:A\to B$ is a way of concisely writing ``$f$ is a function with domain $A$ and
codomain $B$.''

@@

>>>>

Lastly, we define the {\it symmetric difference} $A \symdiff B$ of two sets $A$ and $B$ to be the set of all
elements that are either in $A$ or in $B$ {\it but not both}.
Invoking the above example one last time, we
have $A\symdiff B = \{2,3,4,5\}$.
We can express it as using unions, intersections,
and complements as
\edef\eqsymdiffdef{\the\eqcount}
$$A \symdiff B = (A\cup B) \cap \bar{A \cap B}.\adveq$$
To practise using all the different operations we just learned,
convince yourself that the following are three more valid ways to express the symmetric difference:
$$A\symdiff B = (A\cup B)\setminus (A\cap B) = (A\setminus B) \cup (B\setminus A) =
\bar{\bar{A\cup B} \cup (A\cap B)}$$
More set identities abound. We state the following proposition without proof; you should try going
though this list and convincing yourself that each identity holds, for all sets $A$, $B$, and $C$.
(This is a great way of practising proofs by double inclusion.)

@@

These laws have names, some of which we will use more often than others:
(i) is called the identity law, (ii) the domination law, (iii) the idempotent law,
(iv) the law of double negation, (v) the commutative law, (vi) the associative law,
(vii) the distributive law, (viii) the absorption law, and (ix) the complement law.

@@

\medskip\boldlabel\datestamp{03}{ix} De Morgan's laws.
There are two important laws relating complements with union and intersection.
We shall state them as a proposition, this time giving a proof (of one of them).

@@

To actually prove that $2^{A\cup B} \subseteq 2^A \cup 2^B$ doesn't hold in general, we need to find
a {\it counterexample}. That is, we need to construct sets $A$ and $B$ such that the statement is false.
In this case, we can let $A = \{1,2\}$, $B = \{3,4\}$, so that $A\cup B = \{1,2,3,4\}$.
Then the set $\{1,3\}$ is a subset of $A\cup B$ but is not a subset of $A$ and it is not a subset of $B$.
In other words, $\{1,3\}\in 2^{A\cup B}$ but $\{1,3\} \notin 2^A\cup 2^B$, proving that
$2^{A\cup B} \not\subseteq 2^A \cup 2^B$.
\bye

@@

A {\it function} $f$ from a set $A$ to a set $B$ is a subset $f\subseteq A\times B$ such that
for every $a\in A$, there is exactly one $b\in B$ such that $(a,b)\in f$. (If there is no $b\in B$,
or more than one, we say that $f$ is not well-defined.
If $(a,b)\in f$,
we write $f(a) = b$. The set $A$ is called the {\it domain} and the set $B$ is the {\it codomain}.
The notation $f:A\to B$ is a way of concisely writing ``$f$ is a function with domain $A$ and
codomain $B$.''

@@

So we can find subset of $\NN$ equipotent with it. It turns out we can also find supersets of $\NN$ with
the same property.

\proclaim Theorem \advthm. We have $|\NN| = |\ZZ|$.

\proof We define $f : \NN\to \ZZ$ by
$$f(n) = \cases{{n\over 2}, & if $n$ is even;\cr {-{n+1\over 2}}, & if $n$ is odd.}$$
We shall show that $f$ is a bijection.

To show that $f$ is injective, there are four cases. Note first that if $n$ is even, then $f(n)\ge 0$,
and if $n$ is odd, then $f(n) < 0$. So if $f(m) = f(n)$ for some $m,n\in \NN$, then $f(m)$ and $f(n)$ must
either both be negative, or both be nonnegative. Either way, $m$ and $n$
are either both even or they are both odd. If $m$ and $n$ are both even, then from $f(m) = f(n)$ we derive
$${m\over 2} = {n\over 2},$$
whence multiplying by $2$ on both sides we see that $m=n$. If $m$ and $n$ are both odd, then
$$-{m+1\over 2} = -{n+1\over 2},$$
so, multiplying by $-2$ and subtracting $1$ from both sides we have $m=n$ in this case as well.

Now we show that $f$ is surjective. Let $k\in \ZZ$. If $k\ge 0$, then consider $n=2k$. We have
$$f(n) = f(2k) = {2k\over 2} = k.$$
If $k<0$, then consider $n=-2k-1$. (Check that this is an element of $\NN$.) Then
$$f(n) = f(-2k-1) = -{-2k-1+1\over 2} = -{-2k\over 2} = k.$$
This shows that $f$ is surjective, so $f$ is in fact bijective and we conclude that $\NN=\ZZ$.\slug

We say that a set $A$ is {\it countably infinite} if there exists a bijection $f:\NN\to A$, that is,
if $|\NN| = |A|$.
A set is said to be {\it countable} if it is either finite or countably infinite.
Otherwise it is called {\it uncountable}.
The previous theorem shows that $\ZZ$ is countably infinite.
If a set $A$ is countable, then there exists an enumeration
$$A = \{a_0, a_1, a_2, \ldots\}.$$
This is useful in the proof of the following theorem.

\proclaim Theorem \advthm. If $A$ and $B$ are countably infinite sets,
then $A\times B$ is also countably infinite.

\noindent{\it Proof sketch.}\enspace
Enumerate $A = \{a_1, a_2, \ldots\}$ and $B = \{b_1, b_2, \ldots\}$. Now given
$(a_i, b_j)\in A\times B$ for some $i,j\in \NN$, we can let
$$f(a_i, b_j) = {(i+j+1)(i+j+2)\over 2} - i - 1.$$
Work out the values of $f(a_i, b_j)$ for small choices of $i$ and $j$, and draw a picture to convince yourself that
$f : A\times B \to\NN$ is a bijection. (Rigorously proving this fact
is quite tedious, so we omit the details.) This means that $f^{-1}$
is a bijection from $\NN$ to $A\times B$, so $A\times B$ is countably infinite.\slug

\proclaim Corollary \advthm. We have $|\ZZ\times \ZZ| = |\NN|$.

\proof
We proved earlier that $\ZZ$ is countably infinite, so we may apply the previous theorem with $A = B = \ZZ$.\slug

One can also show that $|\QQ| = |\NN|$. We will not include the proof, but the idea is that there
is a surjective map from $\ZZ\times\bigl( \ZZ \setminus \{0\}\bigr)\to \QQ$ given by $(m,n)\mapsto m/n$.
On the other hand, there are infinite sets that are not countably infinite. An {\it infinite binary
string} is a function $f:\NN\to \{0,1\}$. (By listing out
$$ f(0), f(1), f(2), \ldots$$
you get an infinite string of $0$s and $1$.)

\proclaim The set $A$ of all infinite binary strings is uncountable.

\proof Certainly $A$ is not finite, so we need to show that $|\NN|\ne A$.

Let $f : \NN\to A$. We shall show that $f$ is not surjective. For all $m,n\in \NN$, let $a_{m,n}$
be the $n$th bit of $f(m)$. Consider the infinite binary string
$$s = (1-a_{0,0},1-a_{1,1}, 1-a_{2,2}, 1-a_{3,3}, \ldots).$$
This string $s$ cannot equal $f(m)$ for any $m\in \NN$, since given an arbitrary $m$,
the $m$th bit of $f(m)$ is $a_{m,m}$, whereas the $m$th bit of $s$ is $1-a{m,m}$. Hence there is some
$s\in A$ such that $f(m)\ne s$ for all $m\in \NN$. So $f$ is not surjective.

Since there can be no surjection $f:\NN\to A$, {\it a fortiori} we cannot have a bijection $\NN\to A$.
We conclude that $|A| \ne |\NN|$.\slug

The set of all infinite binary strings is in bijection with the set of all real numbers in the interval
$[0,1)$. If we're being extra pedantic, we need to forbid an infinite trailing string of all $1$s,
since, e.g., $0.0111\ldots = 0.1000\ldots$ after carrying the $1$s (akin to the fact
that, e.g., $0.0999\ldots = 0.1$). After dealing with this detail, we will have shown that
$\bigl| [0,1]\bigr| \ne |\NN|$.
